Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists

### 2024-10-18 paper.Rmd
updates to paper

---
title: "Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists"
author:
  - name: Ronald M. Visser
    orcid_id: 0000-0001-6966-1729
    roles: [writing, editing, analysis, visualization]
    affiliation: Saxion University of Applied Sciences
  - name: Karsten Lambers 
    roles: [writing, editing]
    orcid_id: 0000-0001-6432-0925
    affiliation: Leiden University
  - name: Kenneth Aitchison 
    roles: [writing, editing, software, project administration]
    orcid_id: 0000-0001-6594-1408
    affiliation: Landward Research Teoranta
  - name: Annemarie Jutte
    orcid_id: 0009-0004-9322-0674
    roles: [writing, editing, software]
    affiliation: Saxion University of Applied Sciences
  - name: Doug Rocks-Macqueen 
    orcid_id: 0000-0001-9021-0260
    roles: [writing, editing, software]
    affiliation: Landward Research Teoranta
  - name: Laura van der Knaap 
    roles: [writing, editing, software]
    orcid_id: 0000
    affiliation: Leiden University
  - name: Iza Romanowska 
    orcid_id: 0000-0002-9487-2111
    roles: [writing, editing]
    affiliation: Aarhus University
  - name: Tom Brughmans
    roles: [writing, editing]
    orcid_id: 0000-0002-1589-7768
    affiliation: Aarhus University
  - name: Adam Pažout
    roles: [writing, editing]
    orcid_id: 0000-0001-7745-5634
    affiliation: Aarhus University
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  distill::distill_article:
    number_sections: yes
    fig_caption: yes
  bookdown::pdf_document2:
    number_sections: yes
    keep_tex: true
    fig_caption: yes
  bookdown::word_document2:
    number_sections: yes
    fig_caption: yes
bibliography: references.bib
csl: journal-of-computer-applications-in-archaeology.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
---

@@

@@

In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research. This tutorial consists of 4 lessons. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.

![Screenhot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al. (2002).](paper_files/Screenshot_ABMA_tutorial_1_lesson_3.png)

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials. Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

>>>>

---
title: "Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists"
authors:
  - name: Ronald M. Visser
    orcid_id: 0000-0001-6966-1729
    roles: [writing, editing, analysis, visualization]
    affiliation: Saxion University of Applied Sciences
  - name: Karsten Lambers 
    roles: [writing, editing]
    orcid_id: 0000-0001-6432-0925
    affiliation: Leiden University
  - name: Kenneth Aitchison 
    roles: [writing, editing, software, project administration]
    orcid_id: 0000-0001-6594-1408
    affiliation: Landward Research Teoranta
  - name: Annemarie Jutte
    orcid_id: 0009-0004-9322-0674
    roles: [writing, editing, software]
    affiliation: Saxion University of Applied Sciences
  - name: Doug Rocks-Macqueen 
    orcid_id: 0000-0001-9021-0260
    roles: [writing, editing, software]
    affiliation: Landward Research Teoranta
  - name: Laura van der Knaap 
    roles: [writing, editing, software]
    orcid_id: 0000
    affiliation: Leiden University
  - name: Iza Romanowska 
    orcid_id: 0000-0002-9487-2111
    roles: [writing, editing]
    affiliation: Aarhus University
  - name: Tom Brughmans
    roles: [writing, editing]
    orcid_id: 0000-0002-1589-7768
    affiliation: Aarhus University
  - name: Jeroen Linssen
    orcid_id: 0000-0002-2626-1837
    roles: [writing, editing]
    affiliation: Saxion University of Applied Sciences
  - name: Adam Pažout
    roles: [writing, editing]
    orcid_id: 0000-0001-7745-5634
    affiliation: Aarhus University
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  distill::distill_article:
    number_sections: yes
    fig_caption: yes
  bookdown::pdf_document2:
    number_sections: yes
    keep_tex: true
    fig_caption: yes
  bookdown::word_document2:
    number_sections: yes
    fig_caption: yes
bibliography: references.bib
csl: journal-of-computer-applications-in-archaeology.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
---

@@

**Abstract**

Agent-based modelling (ABM) is one of the most applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists. However, the access to vocational training for ABM is limited. In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation. For this purpose we created online and open tutorials for ABM for archaeologists. In addition, how-to guides and support material was created, both for teachers and students. The material adheres to international (European) standards in relation to education and the Digital Skills Passport for archaeology. The teaching material is well tested using a diverse group of archaeologists during various international conferences and meetings. It was shown that the knowledge and skills of the participants in relation to ABM improved. In addition, the demand for the teaching material and the workshops was high, ensuring that the tutorials on ABM will be used in the future.

@@

In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research (\@ref(fig:screenshot-lesson-3)). This tutorial consists of 4 lessons. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.

```{r, screenshot-lesson-3, echo = FALSE, out.width="100%", fig.cap = "Screenshot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al. (2002)."}
knitr::include_graphics("paper_files/ScreenshotABMAtutorial1lesson3.png")
```

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials \@ref(fig:workshop-impression). Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

```{r workshop-impression, echo = FALSE, out.width="100%", fig.cap = "Impression of various workshops (CAA 2023 in Amsterdam and EAA 2023 in Belfast) with participants working."}
knitr::include_graphics("paper_files/CollageWorkshops.jpg")
```

### 2024-10-18 paper.tex
correction
updates to paper

\title{Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists}
\author{true \and true \and true \and true \and true \and true \and true \and true \and true}
\date{September 18, 2024}

@@

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research. This tutorial consists of 4 lessons. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.

\begin{figure}
\centering
\includegraphics{paper_files/Screenshot_ABMA_tutorial_1_lesson_3.png}
\caption{Screenhot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al.~(2002).}
\end{figure}

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials. Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-aalpoel2024}{}}%
Aalpoel, R, Baan, E, Lucas, S, Mulder, J, Regts, S, Sijbrant, A, Jutte, A and Visser, R. 2024 \emph{Teaching material Agent-Based Modelling for Archaeologists (EN/NL)} DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10950153}{10.5281/zenodo.10950153}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-visser2024}{}}%
Visser, R, Lambers, K, Aitchison, K, Jutte, A, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I and Brughmans, T. 2024 \emph{The Digital Competence Framework for Citizens (2.2) and Agent-Based modelling for Archaeologists}. DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.11046488}{10.5281/zenodo.11046488}.

>>>>

\title{Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists}
\author{}
\date{\vspace{-2.5em}October 18, 2024}

@@

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\textbf{Abstract}

Agent-based modelling (ABM) is one of the most applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists. However, the access to vocational training for ABM is limited. In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation. For this purpose we created online and open tutorials for ABM for archaeologists. In addition, how-to guides and support material was created, both for teachers and students. The material adheres to international (European) standards in relation to education and the Digital Skills Passport for archaeology. The teaching material is well tested using a diverse group of archaeologists during various international conferences and meetings. It was shown that the knowledge and skills of the participants in relation to ABM improved. In addition, the demand for the teaching material and the workshops was high, ensuring that the tutorials on ABM will be used in the future.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research (\ref{fig:screenshot-lesson-3}). This tutorial consists of 4 lessons. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/ScreenshotABMAtutorial1lesson3} \caption{Screenshot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al. (2002).}\label{fig:screenshot-lesson-3}
\end{figure}

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials \ref{fig:workshop-impression}. Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/CollageWorkshops} \caption{Impression of various workshops (CAA 2023 in Amsterdam and EAA 2023 in Belfast) with participants working.}\label{fig:workshop-impression}
\end{figure}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-aalpoel2024}{}}%
Aalpoel, R, Baan, E, Lucas, S, Mulder, J, Regts, S, Sijbrant, A, Jutte, A and Visser, RM. 2024 \emph{Teaching material Agent-Based Modelling for Archaeologists (EN/NL)} DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10950153}{10.5281/zenodo.10950153}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-visser2024}{}}%
Visser, RM, Lambers, K, Aitchison, K, Jutte, A, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I and Brughmans, T. 2024 \emph{The Digital Competence Framework for Citizens (2.2) and Agent-Based modelling for Archaeologists}. DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.11046488}{10.5281/zenodo.11046488}.
